---
title: "Geographic Variation in Opioid Mortality by Race/Ethnicity, 1999--2016"
subtitle: "Identifying epidemic hotspots"
author: 
  - "Mathew Kiang\\inst{1}"
  - "Monica Alexander\\inst{2}"
  - "Zhe Zhang\\inst{3}"
  - "Jarvis Chen\\inst{1}"
institute:
  - "\\inst{1}Deptartment of Social and Behavioral Sciences \\newline  Harvard TH Chan School of Public Health"
  - "\\inst{2}Department of Demography \\newline University of California, Berkeley"
  - "\\inst{3}Heinz College \\newline Carnegie Mellon University"
# date: "`r format(Sys.time(), '%d %B %Y')`"
date: "PAA 2018: Denver, CO"
# fontsize: 12pt
classoption: 
  - compress
  # - "aspectratio=169"
output: 
  beamer_presentation:
    theme: "CambridgeUS"
    color: "dolphin"
    fonttheme: "structurebold"
    incremental: true
    keep_tex: true
    includes:
      in_header: paa_header.tex
    dev: cairo_pdf
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(here)
library(geofacet)
library(statebins)
library(viridis)
library(gridExtra)
```


## Introduction

- Opioid-related mortality is a massive public health issue
    - Over 42,000 opioid-related deaths in 2016
    - More than deaths due to motor vehicle accidents
    - More than deaths due to firearms
    
- More alarming than the magnitude is the trajectory:
    - Current rate of increase similar to peak of HIV epidemic
    - "Triple-wave" epidemic[^1]
        - Not one, but multiple, overlapping epidemics
        - Variation by type of opioid


[^1]: For examples, see: Unick *et al.* 2013; Compton *et al.* 2016; Ciccarone 2017

## Increased over 2,500% since 1980

```{r, cache=TRUE}
f1 <- readRDS(here("plots", "grobs", "presentation_fig1.RDS"))
print(f1)
```

## Wave 1: Prescription Opioids, 1990--2010

```{r, cache=TRUE}
f2 <- readRDS(here("plots", "grobs", "presentation_fig2.RDS"))
print(f2)
```

## Wave 2: Heroin, 2010--current

```{r, cache=TRUE}
f3 <- readRDS(here("plots", "grobs", "presentation_fig3.RDS"))
print(f3)
```

## Wave 3: Synthetic Opioids, 2013--current

```{r, cache=TRUE}
f4 <- readRDS(here("plots", "grobs", "presentation_fig4.RDS"))
print(f4)
```

## Substantial variation: By race/ethnicity
```{r, cache=TRUE}
# f5 <- readRDS(here("plots", "grobs", "presentation_fig5.RDS"))
# print(f5)
```

## Substantial variation: By geography
```{r, cache=TRUE}
# f6 <- readRDS(here("plots", "grobs", "presentation_fig6.RDS"))
# print(f6)
```

## Aims of the paper

1. Systematically describe the opioid epidemic from 1999 to 2016 by state, race/ethnicity, and opioid type.
1. Systematically describe the *current* epidemic.
1. Identify "epidemic hotspots" --- areas with both high mortality rates and rapid increases in mortality.

## Methods

- Multiple cause of death data from NCHS
    - 1999--2015, restricted-access
    - 2016, public-access (pending request to restricted file)
- Age-standardized rates by state, race/ethnicity, and opioid type
    - Underlying cause: X40--X44, X60--X64, X85, Y10--Y14
    - Contributory cause: T40.0--T40.4 and T40.6
- Joinpoint regression (stratified by race/ethnicity, state, opioid type)
    - Segmented weighted least squares regression
    - Grid-search to find optimal joinpoint locations
    - Permutation test to find optimal joinpoint model (0-3 joins)
- Defined statistical significance as $P < 0.01$

## Example Joinpoint Results: Maryland
```{r, cache=TRUE}
# f7 <- readRDS('./plots/grobs/fig_example_jp_MD.RDS')
# print(f7)
```

## Example Joinpoint Results: All opioids
```{r, cache=TRUE, fig.height=6, fig.width=10}
# f8 <- readRDS('./plots/grobs/fig_jp_results_geofacet_opioid.RDS')
# f8 <- f8 + theme(legend.position = 'none')
# print(f8)
```

## Current Mortality
```{r, cache=TRUE, fig.height=6.5, fig.width=11}
# f9 <- readRDS('./plots/grobs/current_rates_map.RDS')
# print(f9)
```

## Current Trajectory
```{r, cache=TRUE, fig.height=6.5, fig.width=11}
# f10 <- readRDS('./plots/grobs/current_trajectory_map.RDS')
# print(f10)
```


## Epidemic Hotspots
```{r, cache=TRUE, fig.height=7.5, fig.width=11, fig.align='center'}
# p_legend <- readRDS('./plots/grobs/hotspots_legend.RDS')
# p1 <- readRDS('./plots/grobs/hotspots_map.RDS')
# 
# grid.arrange(p1, p_legend,  heights = c(4, 1.5), nrow=2)
```

## Results

- There were no decreases in opioid mortality, by type, for either race/ethnicity, anywhere
- Where there were increases, it is just as bad (or worse) for the black population
- Increases are driven by heroin and synthetic opioids
    - Heroin APC: 30-34\% for both populations
    - Synthetic opioid APC: 70\% for whites, 150\% for blacks
    - Synthetic opioids are doubling in 12 states for whites and 18 for blacks
- Strong geographical clustering in heroin and synthetic opioid mortality
    - Little evidence this is due to differences in opioid use
    - Source of heroin/synthetic opioids is important

## Conclusions

- Interventions must be local and tailored to region, race/ethnicity, and opioid type
- Supply-side interventions need to be balanced with harm reduction interventions
    - Syringe exchange programs
    - Medication assisted treatment
    - Increased naloxone access
    - Drug testing on-site and point-of-site
    - Supervised consumption sites
- Surveillance of illicit markets needs to be dramatically improved
    - Better measurement of potency of drugs
    - Types of drugs
    - Cost and availability
    
## Thank you
\Large
\begin{center}
\textbf{Code and interactive results explorer:} \\ \texttt{https://tiny.cc/paa2018} \newline \newline

 mathewkiang.com | \faGithub: mkiang | \faTwitter: @mathewkiang 

\end{center}


## Epidemic Hotspots
```{r, cache=TRUE, fig.height=5, fig.width=8}
# f14 <- readRDS('./plots/grobs/hotspots_scatter_final.RDS')
# print(f14)
```

